---
title: "Analytics Assignment 3 – Airbnb Cape Town"
output: html_document
date: "2026-05-11"
---

```{r libraries}
library(tidyverse)
library(corrplot)
library(FactoMineR)
library(factoextra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cluster)
library(kohonen)
library(mclust)
library(NbClust)
library(sf)
library(tmap)

library(maptiles)
library(fpc)
```

# Exploratory Data Analysis

Notes about the data:

-   The Number of actual listings for a host is a subset of the "host listing count" column and itself a subset of "total host listing count"
-   Wards are geographical divisions for electoral purposes. Neighborhood is like areas like Muizenburg. However, the data doesn't that good. A lot of just Cape Town which doesn't mean much -   Booked, Revenue, Availability (time_frame: 30, 60, ... 365)
-   Number of reviews. ltm = 12 months

Quick Data processing:
-   Rename variables
-   Discard id, picture url, host id (can be inferred from host name)
-   Make categorical variables factors

```{r dataLoading}
load("airbnbListings.RData")
# str(Airbnb_Excel_Lab_AAMB_Co)
head(Airbnb_Excel_Lab_AAMB_Co)

airbnb_data <- Airbnb_Excel_Lab_AAMB_Co %>%
  rename(
    Host_Name     = `Host_Name_Trimmed & Proper`,
    Neighbourhood = `Neighbourhood_Proper & Trim`,
    Wards         = `Neighbourhood_Cleansed`,
    Property_Type = `Property_Type_Proper & Trim`,
    Room_Type     = `Room_Type_Proper & Trim`
  ) %>%
  select(-ID, -Picture_URL, -Host_ID) %>%
  mutate(
    across(c(
      Room_Type,
      Property_Type,
      Neighbourhood,
      Wards
    ), as.factor),
  )

head(airbnb_data)
# str(airbnb_data) # you can check same info in environment panel
# summary(airbnb_data)
```

## NA values and Data Imputation

NA variables:

-   Neighborhood has most NA values. We can impute common neighborhood in assigned ward.
-   Bedrooms, Beds, Bathrooms. We can impute with Median
-   Review_Score: Accuracy, Cleanliness, Checkin, Communication, Location, Value. We can impute with Median

<!-- Potential ideas: -->
<!-- -   Have a occupancy variable (booked/30) -->
<!-- -   Take the log of price -->

```{r naData}
# check of NA values
airbnb_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))
```



```{r dataImputation}
# beds, bathroom, bedroom ward level imputation
airbnb_data <- airbnb_data %>%
  group_by(Wards) %>%
  mutate(across(
    c(Bedrooms, Beds, Bathrooms),
    ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)
  )) %>%
  ungroup() # Always ungroup!

# Replace any remaining NAs with global median.
airbnb_data <- airbnb_data %>%
  mutate(across(
    c(Bedrooms, Beds, Bathrooms),
    ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)
  ))


# Review: Ward level imputation
airbnb_data <- airbnb_data %>%
  group_by(Wards) %>%
  mutate(across(
    starts_with("Review_Scores"),
    ~ ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x)
  )) %>%
  ungroup()

# Replace any remaining NAs with global median.
airbnb_data <- airbnb_data %>%
  mutate(across(
    starts_with("Review_Scores"),
    ~ ifelse(is.nan(.x) | is.na(.x), mean(.x, na.rm = TRUE), .x)
  ))

```

```{r}
colSums(is.na(airbnb_data))
```


## Outliears and Anomolies

-   Neighborhood has many generic labels and has redundant information captured by wards. Dropping the column seems to be a good idea
-   Price seems to have a heavy tail expensive prices.
- Looking at overlap table of property type and room type. Property type is just a fine grained representation of Room Type.
- Outliers in Bathrooms, Bedrooms with 50?? Beds with 70??
- Availability is directly -1 correlation to booking. Hence, dropping this seems to be also a good idea.



```{r}
neighbourhood_table <- as.data.frame(table(airbnb_data$Neighbourhood)) %>% arrange(desc(Freq))
neighbourhood_table

ward_table <- as.data.frame(table(airbnb_data$Wards)) %>% arrange(desc(Freq))
ward_table

```


```{r}
airbnb_data %>% count(Room_Type, sort = TRUE)
airbnb_data %>% count(Property_Type, sort = TRUE)

room_property_table <- table(airbnb_data$Room_Type, airbnb_data$Property_Type)

# View the table
print(room_property_table)
```


```{r}
host_vars <- c("Host_Listings_Count", "Host_Total_Listings_Count")
physical_vars <- c("Accommodates", "Bathrooms", "Bedrooms", "Beds")
pricing_vars <- c("Price")
financial_vars <- c("Revenue_30", "Revenue_60", "Revenue_90", "Revenue_365")
booking_vars <- c("Booked_30", "Booked_60", "Booked_90", "Booked_365")
availability_vars <- c(
  "Availability_30",
  "Availability_60",
  "Availability_90",
  "Availability_365"
)
review_count_vars <- c(
  "Number_of_reviews",
  "Number_of_Reviews_ltm",
  "Number_of_Reviews_l30d",
  "Reviews_per_Month"
)
review_score_vars <- c(
  "Review_Scores_Rating",
  "Review_Scores_Accuracy",
  "Review_Scores_Cleanliness",
  "Review_Scores_Checkin",
  "Review_Scores_Communication",
  "Review_Scores_Value"
)

plot_hist_group <- function(data, var_list, title, fill_color = "steelblue", bins = 30) {
  data %>%
    select(all_of(var_list)) %>%
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
    ggplot(aes(x = value)) +
    geom_histogram(bins = bins, fill = fill_color, color = "white") +
    facet_wrap(~variable, scales = "free") +
    theme_minimal() +
    labs(title = title, x = "Value", y = "Frequency")
}

# Host Activity
plot_hist_group(airbnb_data, host_vars, "Host Listings Distribution", "darkblue")

# Physical Attributes
plot_hist_group(airbnb_data, physical_vars, "Space and Capacity Distribution", "forestgreen", bins = 15)


# Price
plot_hist_group(airbnb_data, c("Price"), "Price Distribution", "steelblue", bins=60)

# Revenue
plot_hist_group(airbnb_data, financial_vars, "Estimated Revenue Distributions", "indianred")


# Bookings
plot_hist_group(airbnb_data, booking_vars, "Booking Counts (Demand)", "purple")

# Availability
plot_hist_group(airbnb_data, availability_vars, "Listing Availability (Supply)", "darkorchid")

# Review Engagement
plot_hist_group(airbnb_data, review_count_vars, "Review Frequency and Engagement", "darkslategrey")

# Review Scores
plot_hist_group(airbnb_data, review_score_vars, "Detailed Review Scores", "gold", bins = 20)
```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)

df_subset <- get_data_subset()


p_bath <- ggplot(df_subset, aes(x = factor(Bathrooms))) +
  geom_bar(fill = "forestgreen", color = "white", width = 0.9) +
  theme_minimal() +
  labs(title = "Bathrooms", x = NULL, y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

p_bed <- ggplot(df_subset, aes(x = factor(Bedrooms))) +
  geom_bar(fill = "forestgreen", color = "white", width = 0.9) +
  theme_minimal() +
  labs(title = "Bedrooms", x = NULL, y = NULL) +
  theme(plot.title = element_text(hjust = 0.5))

p_beds <- ggplot(df_subset, aes(x = factor(Beds))) +
  geom_bar(fill = "forestgreen", color = "white", width = 0.9) +
  theme_minimal() +
  labs(title = "Beds", x = "Units", y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

p_price <- ggplot(df_subset, aes(x = Log_Price)) +
  geom_histogram(bins = 30, fill = "forestgreen", color = "white") +
  theme_minimal() +
  labs(title = "Log_Price", x = "Units", y = NULL) +
  theme(plot.title = element_text(hjust = 0.5))

plot_transformed <- (p_bath + p_bed) / (p_beds + p_price)


p_bath_untrans <- ggplot(airbnb_data, aes(x = Bathrooms)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Bathrooms", x = NULL, y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

p_bed_untrans <- ggplot(airbnb_data, aes(x = Bedrooms)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Bedrooms", x = NULL, y = NULL) +
  theme(plot.title = element_text(hjust = 0.5))

p_beds_untrans <- ggplot(airbnb_data, aes(x = Beds)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Beds", x = "Units", y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))

p_price_untrans <- ggplot(airbnb_data, aes(x = Price)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Price", x = "Units", y = NULL) +
  theme(plot.title = element_text(hjust = 0.5))

plot_untransformed <- (p_bath_untrans + p_bed_untrans) / (p_beds_untrans + p_price_untrans)

plot_untransformed | plot_transformed
```


```{r}
# check the outliers for beds, bedrooms, and bathrooms
airbnb_data %>%
  select(Bedrooms, Bathrooms, Beds) %>%
  # Remove NAs to avoid plot errors
  drop_na() %>%
  pivot_longer(cols = everything(),
               names_to = "Feature",
               values_to = "Count") %>%
  ggplot(aes(x = Feature, y = Count, fill = Feature)) +
  # Create the box plot
  geom_boxplot(outlier.color = "red", outlier.shape = 1, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Box Plot of Physical Capacity Features",
       subtitle = "Full dataset (N = 16,319)",
       x = "Variable",
       y = "Count") +
  scale_fill_brewer(palette = "Pastel1") +
  theme(legend.position = "none")
```


## Correlation

```{r correlation}

cor_all <- airbnb_data %>%
  select(where(is.numeric)) %>%
  select(
    -Host_Listings_Count,
    -Host_Total_Listings_Count,
    -Latitude,
    -Longitude,
    -starts_with("Booked"),
    -starts_with("Availability"),
    -starts_with("Revenue")
   ) %>%
  cor(use = "complete.obs")


corrplot(cor_all,
         method = "color",
         type = "upper",
         tl.cex = 0.45,    
         tl.pos = "l",
         tl.col = "black",
         addCoef.col = "black",
         number.cex = 0.35,    
         mar = c(0, 0, 1, 0)
         # title = "Correlation Matrix of Numeric Features"
         )


cor_booking_avail <- airbnb_data %>%
  select(starts_with("Booked"), starts_with("Availability")) %>%
  cor(use = "complete.obs")

corrplot(cor_booking_avail,
         method = "color",
         type = "upper",
         tl.cex = 0.45,        
         # tl.pos = "l",
         tl.col = "black",
         addCoef.col = "black",
         number.cex = 0.35,     
         mar = c(0, 0, 1, 0),
         title = "Correlation Matrix")

cor_reviews <- airbnb_data %>%
  select(contains("review")) %>%
  cor(use = "complete.obs")

corrplot(cor_reviews,
         method = "color",
         type = "upper",
         tl.cex = 0.45,         
         # tl.pos = "l",
         tl.col = "black",
         addCoef.col = "black",
         number.cex = 0.35,     
         mar = c(0, 0, 1, 0),
         title = "Correlation Matrix")


cor_all_minus_review <- airbnb_data %>%
  select(where(is.numeric)) %>%
  select(
    # -Host_Listings_Count,
    # -Host_Total_Listings_Count,
    -Latitude,
    -Longitude,
    # -starts_with("Booked"),
    -starts_with("Availability"),
    -contains("review")
   ) %>%
  cor(use = "complete.obs")


corrplot(cor_all_minus_review,
         method = "color",
         type = "upper",
         tl.cex = 0.45,      
         # tl.pos = "l",
         tl.col = "black",
         addCoef.col = "black",
         number.cex = 0.35,     
         mar = c(0, 0, 1, 0),
         title = "Correlation Matrix")

```

## PCA

We split PCA into two groups from findings from the before correlation matrix
-   Review variables. We seem to get 2 PCA.
-   Everything. We seem to get 4 PCA

Other PCA:
-   Price and related things with price (bathroom, size,..). Financial indicators
-   Availability. Short term indicators
-   House features. Physical capacity
-   Year metrics of booking and revenue. Long term indicators

Reviews PCA:
-   Number of reviews. Review Traffic
-   Review Scores. Guest Satisfaction

```{r}
pca_prep <- airbnb_data %>%
  select(where(is.numeric)) %>%
  select(
    -Host_Listings_Count,
    -Host_Total_Listings_Count,
    -Latitude,
    -Latitude,
    -Longitude,
    -starts_with("Availability"),
    # -Price
    # starts_with("Booked"),
    # -contains("review")
   ) %>%
  select(contains("review"), Accommodates, Bathrooms, Bedrooms, Beds)


pca_group <- PCA(pca_prep %>% select(-contains("review")), scale.unit = TRUE)
pca_review  <- PCA(pca_prep %>% select(contains("review")), scale.unit = TRUE)

fviz_eig(pca_group, addlabels = TRUE, ylim = c(0, 50))
fviz_eig(pca_review, addlabels = TRUE, ylim = c(0, 50))
```

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
library(FactoMineR) 
library(factoextra) 

pca_prep <- airbnb_data %>%
  select(where(is.numeric)) %>%
  select(
    -Host_Listings_Count,
    -Host_Total_Listings_Count,
    -Latitude,
    -Longitude,
    -starts_with("Availability")
  ) %>%
  select(contains("review"), Accommodates, Bathrooms, Bedrooms, Beds) %>%
  mutate(
    # Log_Price = log(Price),
    # Cap Bedrooms at 5, Bathrooms at 4, Beds at 6, Accommodates at 10
    Bedrooms = pmin(Bedrooms, 5),
    Bathrooms = pmin(Bathrooms,4),
    Beds = pmin(Beds, 6),
    Accommodates = pmin(Accommodates, 10)
  )

pca_group  <- PCA(pca_prep %>% select(-contains("review")), scale.unit = TRUE, graph = FALSE)
pca_review <- PCA(pca_prep %>% select(contains("review")), scale.unit = TRUE, graph = FALSE)


plot_group <- fviz_eig(pca_group, addlabels = TRUE) +
  labs(title = "Scree Plot: Physical Features") +
  theme(plot.title = element_text(hjust = 0.5))

plot_review <- fviz_eig(pca_review, addlabels = TRUE) +
  labs(title = "Scree Plot: Review Scores") +
  theme(plot.title = element_text(hjust = 0.5))



physical_group <- fviz_contrib(pca_group, choice = "var", axes = 1, top = 10, title = "Physical Capacity PC1 Contributions")

review_pca1 <- fviz_contrib(pca_review, choice = "var", axes = 1, top = 10, title = "Review PC1 Contributions")
review_pca2 <- fviz_contrib(pca_review, choice = "var", axes = 2, top = 10, title = "Review PC2 Contributions")

(plot_group | plot_review) / (review_pca1 + review_pca2)
```



```{r}
# PCA Plot 1
fviz_contrib(pca_group, choice = "var", axes = 1, top = 10, title = "Group PC1 Contributions")
fviz_contrib(pca_group, choice = "var", axes = 2, top = 10, title = "Group PC2 Contributions")
fviz_contrib(pca_group, choice = "var", axes = 3, top = 10, title = "Group PC3 Contributions")
fviz_contrib(pca_group, choice = "var", axes = 4, top = 10, title = "Group PC4 Contributions")

# PCA Plot 2
fviz_contrib(pca_review, choice = "var", axes = 1, top = 10, title = "Review PC1 Contributions")
fviz_contrib(pca_review, choice = "var", axes = 2, top = 10, title = "Review PC2 Contributions")
```

## Putting it all together

```{r}

# # Replace numerical variables with both PCA analyses
pca_features <- data.frame(
  Physical_Capacity   = pca_group$ind$coord[, 1],
  Review_Quality      = pca_review$ind$coord[, 1],
  Review_Volume       = pca_review$ind$coord[, 2]
)


# ward averages
ward_averages <- airbnb_data %>%
  group_by(Wards) %>%
  mutate(
    Log_Price = log(Price),
  ) %>%
  summarise(
    Avg_Price     = mean(Log_Price, na.rm = TRUE),
    Avg_Latitude  = mean(Latitude, na.rm = TRUE),
    Avg_Longitude = mean(Longitude, na.rm = TRUE),
    # Listing_Count = n()
  ) %>%
  ungroup()

print(ward_averages)
```


```{r}
# makes price to be log
# add pca features
# add ward level averages
get_data_subset <- function() {
  data <- airbnb_data %>%
  mutate(
    Log_Price = log(Price),
  ) %>%
  select(Wards, Log_Price, Room_Type) %>%
  bind_cols(pca_features) %>%
  left_join(ward_averages, by = "Wards") %>%
  select(-Wards)
}

temp_df <- get_data_subset()
head(temp_df)

```

## Choosing K
```{r date_prep}

set.seed(123)
nstart <- 25

data <- get_data_subset()
data_num <- data %>%
  select(Log_Price, Physical_Capacity, Review_Quality, Review_Volume ) %>%
  as.matrix() %>%
  scale()

# data_cat <- model.matrix(~ Room_Type - 1, data = data)

# data <-cbind(data_num, data_cat)
data <- data_num
```


```{r choosing_k_wss, cache=TRUE}
fviz_nbclust(data, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method (WSS) for K-means")
```

```{r choosing_k_sil}
fviz_nbclust(data, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Average Silhouette Width for K-means")
```

```{r choosing_k_gap}
set.seed(123)
# gap_stat <- clusGap(data, FUNcluster = kmeans, K.max = 10, B = 25, nstart = nstart)
save(gap_stat, file="gap_stat.RData")
load("gap_stat.RData")
plot(gap_stat, main = "Gap Statistic for K-means")
```


```{r choosing_k_nbclust}

set.seed(123)
# res.nb <- NbClust(nhc_data, distance = "euclidean",
#                   min.nc = 2, max.nc = 10,
#                   method = "kmeans", index = "all")
# save(res.nb, file = "nbclustResults.RData")
load("nbclustResults.RData")
table(res.nb$Best.nc["Number_clusters", ])
cat("Majority vote K:",
    as.integer(names(which.max(table(res.nb$Best.nc["Number_clusters", ])))), "\n")
fviz_nbclust(res.nb)

 k_vals <- as.numeric(rownames(res.nb$All.index))
valid_indices <- intersect(colnames(res.nb$All.index), colnames(res.nb$Best.nc))
  for (idx in valid_indices)  {
    best_k <- res.nb$Best.nc["Number_clusters", idx]
    plot(k_vals, res.nb$All.index[, idx], type = "b",
         xlab = "Number of clusters k", ylab = "Index value",
         main = paste("Index:", idx))
    if (!is.na(best_k) && best_k > 0)
      abline(v = best_k, lty = 2, col = "red")
  }
```

## Cluster profilling helpers

```{r cluster-profiling-helpers}
profile_vars <- get_data_subset()

cluster_num_vars <- c("Log_Price", "Physical_Capacity", "Review_Quality", "Review_Volume")

summarise_profile <- function(cluster_vec) {
    profile_vars %>%
      mutate(cluster = factor(cluster_vec)) %>%
      group_by(cluster) %>%
      summarise(
        n                      = n(),
        mean_log_price         = round(mean(Log_Price,         na.rm = TRUE), 3),
        mean_physical_capacity = round(mean(Physical_Capacity, na.rm = TRUE), 3),
        mean_review_quality    = round(mean(Review_Quality,    na.rm = TRUE), 3),
        mean_review_volume     = round(mean(Review_Volume,     na.rm = TRUE), 3),
        dominant_room_type     = names(which.max(table(Room_Type)))
      )
  }

  plot_profile <- function(cluster_vec, title) {
    k   <- length(unique(cluster_vec))
    mat <- profile_vars %>%
    select(all_of(cluster_num_vars)) %>%
    scale() %>%
    as.data.frame() %>%
    mutate(cluster = cluster_vec) %>%
    group_by(cluster) %>%
    summarise(across(everything(), mean)) %>%
    select(-cluster) %>%
    as.matrix()

    as.data.frame(mat) %>%
    mutate(cluster = factor(1:k)) %>%
    pivot_longer(-cluster, names_to = "variable", values_to = "mean") %>%
    mutate(variable = factor(variable, levels = cluster_num_vars)) %>%
    ggplot(aes(x = variable, y = mean, colour = cluster, group = cluster)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 1.2) +
    geom_point(size = 3) +
    scale_colour_brewer(palette = "Set1", name = "Cluster") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
    labs(title = title, x = NULL, y = "Standardised Mean")
}

```





## Hierarchical Clustering

```{r}

set.seed(123)
data <- get_data_subset()
data_num <- data %>%
  select(Log_Price, Physical_Capacity, Review_Quality, Review_Volume ) %>%
  as.matrix() %>%
  scale()
data <- data_num

hierarchical_clustering <- hclust(
  daisy(data,metric="gower"),
  method="complete"
)

plot(hierarchical_clustering, labels = F,-1)
rect.hclust(hierarchical_clustering, k =4, border = 2:11)
```

```{r}
hierarchical_clustering_cut <- cutree(hierarchical_clustering, k=4)
 # view cluster allocation
# names(hierarchical_clustering_cut) <- rownames(data)
# sort(hierarchical_clustering_cut)


data_clustered <- as.data.frame(data) %>%
  mutate(Cluster = as.factor(hierarchical_clustering_cut))
```

#proffiling

```{r hc_profile_plot}
print(summarise_profile(hierarchical_clustering_cut))
plot_profile(hierarchical_clustering_cut, "Hierarchical Clustering Profiles (k = 4)")
```

```{r}
# cluster_summary <- data_clustered %>%
#   group_by(Cluster) %>%
#   summarise(
#     Listings_Count = n(),
#     Avg_Log_Price = mean(Log_Price),
#     Avg_Accommodates = mean(Accommodates),
#     Avg_Bedrooms = mean(Bedrooms),
#     Avg_Bathrooms = mean(Bathrooms),
#     # Find the most frequent Room_Type in this cluster
#     Dominant_Room_Type = names(which.max(table(Room_Type)))
#   ) %>%
#   arrange(desc(Listings_Count))
#
# # View the summary table
# print(cluster_summary)
```

```{r}
# Boxplot for Price by Cluster
ggplot(data_clustered, aes(x = Cluster, y = Log_Price, fill = Cluster)) +
  geom_boxplot(show.legend = FALSE) +
  theme_minimal() +
  labs(title = "Distribution of Price across Clusters",
       x = "Cluster ID", y = "Log Price")

# Boxplot for Accommodates by Cluster
ggplot(data_clustered, aes(x = Cluster, y = Accommodates, fill = Cluster)) +
  geom_boxplot(show.legend = FALSE) +
  theme_minimal() +
  labs(title = "Listing Capacity (Accommodates) across Clusters",
       x = "Cluster ID", y = "Guests Accommodated")
```

```{r}
ggplot(data_clustered, aes(x = Cluster, fill = Room_Type)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Room Type Breakdown per Cluster",
       x = "Cluster ID", y = "Percentage of Listings", fill = "Room Type")
```



```{r hc_validation}
hc_dist <- daisy(get_data_subset(), metric = "gower")
sil_hc  <- silhouette(hierarchical_clustering_cut, hc_dist)

cat("HC mean silhouette:", round(mean(sil_hc[, 3]), 3), "\n")

fviz_silhouette(sil_hc) +
  labs(title = "Silhouette Plot — Hierarchical Clustering (k = 3)")
```

## Stability: Bootstrapped ARI

```{r hc_ari}
hc_data <- data
n_hc    <- nrow(hc_data)
B       <- 100
ari_hc_vals <- numeric(B)

set.seed(123)
for (b in 1:B) {
  idx       <- sample(1:n_hc, size = n_hc, replace = TRUE)
  hc_boot   <- hclust(daisy(hc_data[idx, ], metric = "gower"), method = "complete")
  hc_boot_cut <- cutree(hc_boot, k = 10)
  ari_hc_vals[b] <- adjustedRandIndex(hierarchical_clustering_cut[idx], hc_boot_cut)
}

cat("Mean ARI:", round(mean(ari_hc_vals), 3), "\n")
cat("SD ARI:  ", round(sd(ari_hc_vals), 3), "\n")
hist(ari_hc_vals, main = "Bootstrap ARI Distribution (HC)",
     xlab = "ARI", col = "steelblue", breaks = 20)
```

## Stability: Jaccard (clusterboot)

```{r hc_jaccard}
hc_dist_mat <- as.matrix(daisy(data, metric = "gower"))

set.seed(123)
clboot_hc <- clusterboot(
  hc_dist_mat,
  B = 25,
  clustermethod = hclustCBI,
  method = "complete",
  k = 4,
  dissimilarity = TRUE,
  seed = 123
)

cat("Mean Jaccard stability per cluster:\n")
print(round(clboot_hc$bootmean, 3))
```


```{r}

airbnb_map <- airbnb_data %>%
  mutate(
    hierarchical    = factor(hierarchical_clustering_cut),
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
tmap_mode("plot")


tm_shape(airbnb_map) +
  tm_dots(fill = "hierarchical", size = 0.5) +
  tm_layout(title = paste0("Hierarchical Clusters (K = ", 4, ")"))

tiles <- get_tiles(airbnb_map, provider = "OpenStreetMap", zoom = 12)

# hierachical
tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "hierarchical", size = 0.5) +
  tm_layout(title = paste0("Hierarchical Clusters (K = ", 4, ")"))

```


# Non-Hierarchical Clustering

## Data Preparation

```{r nhc_setup}
set.seed(123)
nstart <- 25

data <- get_data_subset()
data_num <- data %>%
  select(Log_Price, Physical_Capacity, Review_Quality, Review_Volume ) %>%
  as.matrix() %>%
  scale()

# data_cat <- model.matrix(~ Room_Type - 1, data = data)

# data <-cbind(data_num, data_cat)
nhc_data <- data_num

```


## Fitting: K-means and CLARA

```{r nhc_fit}
K <- 4

set.seed(123)
km_final   <- kmeans(nhc_data, centers = K, nstart = nstart)
clara_out  <- clara(nhc_data, k = K, samples = 25, pamLike = TRUE)

fviz_cluster(km_final, data = nhc_data, geom = "point",
             ellipse.type = "convex") +
  labs(title = paste0("K-means Clustering (K = ", K, ")"))

fviz_cluster(clara_out, data = nhc_data, geom = "point",
             ellipse.type = "convex") +
  labs(title = paste0("CLARA Clustering (K = ", K, ")"))
```

## Cluster Profiling

```{r nhc_profiling}
print(summarise_profile(km_final$cluster))
print(summarise_profile(clara_out$clustering))
```

```{r nhc_profile_plot}
plot_profile(km_final$cluster,   "K-means Cluster Profiles")
plot_profile(clara_out$clustering, "CLARA Cluster Profiles")
```


## Comparison and Validation

```{r nhc_compare}


# Agreement between the two methods
ari_km_clara <- adjustedRandIndex(km_final$cluster, clara_out$clustering)
cat("ARI (K-means vs CLARA):", round(ari_km_clara, 3), "\n")

# Silhouette for each method
sil_km    <- silhouette(km_final$cluster,    dist(nhc_data))
sil_clara <- silhouette(clara_out$clustering, dist(nhc_data))

cat("K-means mean silhouette:", round(mean(sil_km[, 3]), 3), "\n")
cat("CLARA   mean silhouette:", round(mean(sil_clara[, 3]), 3), "\n")


```

## Stability: Bootstrapped ARI

```{r nhc_ari}

B        <- 100
n        <- nrow(nhc_data)
ari_vals <- numeric(B)

set.seed(123)
for (b in 1:B) {
  idx         <- sample(1:n, size = n, replace = TRUE)
  km_boot     <- kmeans(nhc_data[idx, ], centers = K, nstart = nstart)
  ari_vals[b] <- adjustedRandIndex(km_final$cluster[idx], km_boot$cluster)
}

cat("Mean ARI:", round(mean(ari_vals), 3), "\n")
cat("SD ARI:  ", round(sd(ari_vals), 3), "\n")
hist(ari_vals, main = "Bootstrap ARI Distribution (K-means)",
     xlab = "ARI", col = "steelblue", breaks = 20)
```

## Stability: Jaccard (clusterboot)

```{r nhc_jaccard}
set.seed(123)
clboot_out <- clusterboot(
  nhc_data,
  B = 100,
  clustermethod = kmeansCBI,
  krange = K,
  seed = 123
)

cat("Mean Jaccard stability per cluster:\n")
print(round(clboot_out$bootmean, 3))
```

```{r map_plot}
# plotting maps
airbnb_map <- airbnb_data %>%
  mutate(
    km_cluster    = factor(km_final$cluster),
    clara_cluster = factor(clara_out$clustering)
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
tmap_mode("plot")
# K-means map
tm_shape(airbnb_map) +
  tm_dots(fill = "km_cluster", size = 0.5) +
  tm_layout(title = paste0("K-means Clusters (K = ", K, ")"))

# CLARA map
tm_shape(airbnb_map) +
  tm_dots(fill = "clara_cluster", size = 0.5) +
  tm_layout(title = paste0("CLARA Clusters (K = ", K, ")"))

tiles <- get_tiles(airbnb_map, provider = "OpenStreetMap", zoom = 12)

# K-means
tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "km_cluster", size = 0.5) +
  tm_layout(title = paste0("K-means Clusters (K = ", K, ")"))

# CLARA
tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "clara_cluster", size = 0.5) +
  tm_layout(title = paste0("CLARA Clusters (K = ", K, ")"))


map_kmeans <- tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "km_cluster", size = 0.5) +
  tm_layout(title = paste0("K-means Clusters (K = ", K, ")"), 
            legend.outside = TRUE)

map_clara <- tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "clara_cluster", size = 0.5) +
  tm_layout(title = paste0("CLARA Clusters (K = ", K, ")"), 
            legend.outside = TRUE)

tmap_arrange(map_kmeans, map_clara, ncol = 2)
```

## HC vs NHC Comparison

```{r hc_nhc_compare}
# Evaluate all three solutions on the same Euclidean distance for a fair comparison
sil_hc <- silhouette(hierarchical_clustering_cut, dist(nhc_data))

cat("Mean silhouette scores:\n")
cat("  HC:      ", round(mean(sil_hc[, 3]), 3), "\n")
cat("  K-means: ", round(mean(sil_km[, 3]), 3), "\n")
cat("  CLARA:   ", round(mean(sil_clara[, 3]), 3), "\n")

ari_hc_km    <- adjustedRandIndex(hierarchical_clustering_cut, km_final$cluster)
ari_hc_clara <- adjustedRandIndex(hierarchical_clustering_cut, clara_out$clustering)
cat("\nARI (HC vs K-means):", round(ari_hc_km, 3), "\n")
cat("ARI (HC vs CLARA):  ", round(ari_hc_clara, 3), "\n")
```


```{r}
airbnb_map <- airbnb_data %>%
  mutate(
    SOM    = factor(som_cluster_assignments),
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
tmap_mode("plot")
# K-means map
tm_shape(airbnb_map) +
  tm_dots(fill = "SOM", size = 0.5)
  # tm_layout(title = paste0("SOM"))


tiles <- get_tiles(airbnb_map, provider = "OpenStreetMap", zoom = 12) #chat gave this

# SOM
tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "SOM", size = 0.5)
  # tm_layout(title = paste0("K-means Clusters (K = ", K, ")"))
```


# SOM

```{r}
set.seed(123)
data <- get_data_subset()

data_num <- data %>%
  select(Log_Price, Physical_Capacity, Review_Quality, Review_Volume
         #, Avg_Price
         # Avg_Latitude, Avg_Longitude
         ) %>%

  as.matrix() %>%
  scale()

data_cat <- model.matrix(~ Room_Type - 1, data = data)
som_data <- list(measurements = data_num, room_types = data_cat)


# SOM
som_grid <- somgrid(xdim = 15, ydim = 15, topo = "hexagonal")
# grid_distances <- unit.distances(som_grid)
# start_radius <- quantile(grid_distances, 1/2)

som_model <- supersom(
  data = som_data,
  grid = som_grid,
  rlen = 2000,
  whatmap = c("measurements", "room_types"),
  user.weights = c(1, 0.1),
  alpha= c(0.05, 0.01),
)


# data <- as.data.frame(X_matrix)
# data <- data %>% select(-contains("wards"), -contains("room"), -Longitude, -Latitude)
#
# som_grid <- somgrid(xdim = 20, ydim=20, topo="hexagonal")

# som_model <- som(data_num,grid=som_grid, rlen=1000,
#                  alpha=c(0.05,0.01),
#                  keep.data = TRUE  )
```

```{r}
# 1. Split the plotting device into a 2x2 grid
# mar adjusts margins (bottom, left, top, right) to give the subplots breathing room
png("som_plots_combined.png", width = 1200, height = 1000, res = 120)
par(mfrow = c(2, 2), mar = c(4, 4, 3, 2))

# 2. Call your four plots sequentially
plot(som_model, type = "changes")
plot(som_model, type = "counts")
plot(som_model, type = "dist.neighbours")
plot(som_model, type = "quality")
par(mfrow = c(1, 1))
```

```{r}
maxRed <- function(n, alpha = 1) {
  colorRampPalette(c("#F7F7F7", "#B2182B"), alpha = TRUE)(n)
}

maxGreen <- function(n, alpha = 1) {
  colorRampPalette(c("#F7F7F7", "#1A9850"), alpha = TRUE)(n)
}

maxBlue <- function(n, alpha = 1) {
  colorRampPalette(c("#F7F7F7", "#2166AC"), alpha = TRUE)(n)
}

maxOrange <- function(n, alpha = 1) {
  colorRampPalette(c("#F7F7F7", "orange"), alpha = TRUE)(n)
}

SomModelCodes = som_model$codes[[1]]
# png("som_heatmaps.png", width = 1200, height = 1000, res = 120)
par(mfrow = c(2, 2))

plot(
  som_model,
  type = "property",
  property = SomModelCodes[, 1],
  main = colnames(SomModelCodes)[1],
  palette.name = maxRed
)

plot(
  som_model,
  type = "property",
  property = SomModelCodes[, 2],
  main = colnames(SomModelCodes)[2],
  palette.name = maxGreen
)

plot(
  som_model,
  type = "property",
  property = SomModelCodes[, 3],
  main = colnames(SomModelCodes)[3],
  palette.name = maxBlue
)

plot(
  som_model,
  type = "property",
  property = SomModelCodes[, 4],
  main = colnames(SomModelCodes)[4],
  palette.name = maxOrange
)
```


## Clustering

```{r}
length(som_model$unit.classif)
tail(som_model$unit.classif,15)
table(som_model$unit.classif)

data <- som_model$codes[[1]]
data <- data %>% cbind(som_model$codes[[2]])

set.seed(123)
# find a k
wss <- (nrow(data)-1)*sum(apply(data,2,var))
for (i in 2:15) {
  wss[i] <- sum(kmeans(data, centers=i)$withinss)
}
plot(wss)

fviz_nbclust(data, kmeans, method = "wss") +
  labs(title = "Elbow method for k-means")
```

```{r}
# clustering on the weights of SOM
set.seed(123)
K <- 4
som_cluster <- kmeans(data, centers=K)$cluster
som_cluster_assignments <- som_cluster[som_model$unit.classif]
# plotting som clusters
# plot(som_model,palette.name = rainbow)
# add.cluster.boundaries(som_model, som_cluster)


cluster_cols <- c("grey", "#2166AC", "orange","yellow", "blue", "indianred")
par(mfrow = c(1, 2))
plot(som_model,palette.name = rainbow, bgcol = cluster_cols[som_cluster])
# add.cluster.boundaries(som_model, som_cluster)
```

## Profiling


```{r}
# Tabular Summary Profile
print(summarise_profile(som_cluster_assignments))

# Parallel Coordinate Line Plot Profile
plot_profile(som_cluster_assignments, "SOM Meta-Cluster Profiles")
```

## External Analysis

```{r}
# Silhouette for the SOM clustering
# We use 'data_num' here because it represents the scaled numeric matrix used by the SOM
sil_som <- silhouette(som_cluster_assignments, dist(data_num))
cat("SOM mean silhouette:", round(mean(sil_som[, 3]), 3), "\n")
```

## Cluster Stability

```{r}
B <- 50
n <- nrow(data_num)
som_ari_vals <- numeric(B)

set.seed(123)
for (b in 1:B) {
  idx <- sample(1:n, size = n, replace = TRUE)

  boot_data <- list(
    measurements = data_num[idx, ],
    room_types   = data_cat[idx, ]
  )

  som_boot <- supersom(
    data = boot_data,
    grid = som_grid,
    rlen = 1000,
    whatmap = c("measurements", "room_types"),
    user.weights = c(1, 0.1),
    alpha = c(0.05, 0.01)
  )
  boot_codes <- cbind(som_boot$codes$measurements, som_boot$codes$room_types)
  K <- 3
  boot_node_clusters <- kmeans(boot_codes, centers = K, nstart = 25)$cluster
  boot_assignments <- boot_node_clusters[som_boot$unit.classif]


  som_ari_vals[b] <- adjustedRandIndex(som_cluster_assignments[idx], boot_assignments)
}

cat("SOM Mean Bootstrap ARI:", round(mean(som_ari_vals), 3), "\n")
cat("SOM SD Bootstrap ARI:  ", round(sd(som_ari_vals), 3), "\n")

hist(som_ari_vals, main = "Bootstrap ARI Distribution (SOM)",
     xlab = "ARI", col = "darkorange", breaks = 15)
```

```{r}
airbnb_map <- airbnb_data %>%
  mutate(
    SOM    = factor(som_cluster_assignments),
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
tmap_mode("plot")
# K-means map
tm_shape(airbnb_map) +
  tm_dots(fill = "SOM", size = 0.5)
  # tm_layout(title = paste0("SOM"))


tiles <- get_tiles(airbnb_map, provider = "OpenStreetMap", zoom = 12) #chat gave this

# SOM
tm_shape(tiles) +
  tm_rgb() +
  tm_shape(airbnb_map) +
  tm_dots(fill = "SOM", size = 0.5)
  # tm_layout(title = paste0("K-means Clusters (K = ", K, ")"))

```

